---
title: "My First Closeread Story"
author: "Ian Cheung"
format: 
  closeread-html:
    theme: cosmo
    cr-style:
      section-background-color: white
      narrative-background-color-sidebar: black
      narrative-text-color-sidebar: white
      narrative-font-size: 2em
embed-resources: true
---

```{r setup, include=FALSE}
library(tidyverse)
```

:::: {.cr-section}


We look at how engine size impacts fuel economy using the `mpg` dataset. @cr-plot1


As we look at the relationship between engine displacement and highway miles per gallon, we see larger engines (higher displacement) generally result in lower fuel efficiency. @cr-plot1


::: {#cr-plot1}

```{r}
#| echo: false
#| message: false
#| warning: false

ggplot(data = mpg, aes(x = displ, y = hwy, color = class)) +
  geom_point(size = 3) +
  theme_minimal() +
  labs(title = "Engine Size vs. Highway MPG",
       x = "Engine Displacement (Liters)",
       y = "Highway MPG")
```

:::

As we look under the hood, we can see how mechanical design influences these numbers. @cr-engine-image

::: {#cr-engine-image}
![](https://encrypted-tbn0.gstatic.com/images?q=tbn:ANd9GcTAZpNa3oRz2qe2BO0Ur-9fpGIWs9JVck04Qg&s)
:::

Finally, let's compare the averages across different car classes. @cr-table1

::: {#cr-table1}

```{r}
#| echo: false
#| message: false

mpg |>
  group_by(class) |>
  summarize(Avg_Hwy = round(mean(hwy), 1), Count = n()) |>
  arrange(desc(Avg_Hwy)) |>
  knitr::kable(format = "html")
```

:::
::::