---
title: "Sentiment Analysis"
---

## Introduction

Sentiment analysis examines the emotional tone of text. Rather than just counting words, sentiment analysis assigns emotional values to words and aggregates them to understand the overall emotional trajectory. 

This analysis uses these three approaches:

1. **Dictionary Comparison**: Comparing multiple sentiment lexicons to understand nuances
2. **Sentiment Trajectory**: Tracking how sentiment evolves across the series
3. **Contribution Analysis**: Identifying which words have the greatest impact on sentiment

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(tidytext)
library(plotly)

got <- read_csv("game-of-thrones.csv")
bing <- read_csv("bing.csv")
nrc <- read_csv("nrc.csv")

got_tokens <- got %>%
  unnest_tokens(word, Sentence)

got_tokens <- got_tokens %>%
  group_by(Season, Episode) %>%
  mutate(episode_id = cur_group_id()) %>%
  ungroup()
```

## Part 1: Dictionary Comparison - Bing vs NRC

Different sentiment lexicons capture emotion in different ways. **Bing** provides binary sentiment (positive/negative), while **NRC** provides multiple emotion categories (joy, fear, anger, anticipation, etc.).

### Plot 1: Binary Sentiment Comparison (Bing Lexicon)

The Bing lexicon classifies words as either positive or negative. 

```{r plot1, warning=FALSE}
bing_sentiment <- got_tokens %>%
  inner_join(bing, by = "word") %>%
  count(Season, sentiment) %>%
  pivot_wider(names_from = sentiment, values_from = n, values_fill = 0) %>%
  mutate(net_sentiment = positive - negative,
         Season = as.factor(Season))

bing_plot <- bing_sentiment %>%
  pivot_longer(cols = c(positive, negative), names_to = "sentiment", values_to = "count") %>%
  ggplot(aes(x = Season, y = count, fill = sentiment)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c("positive" = "green", "negative" = "red")) +
  labs(title = "Positive vs Negative Words by Season", x = "Season", y = "Word Count", fill = "Sentiment") +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", size = 12))

ggplotly(bing_plot, tooltip = c("x", "y", "fill"))
```

### Plot 2: Emotion Categories (NRC Lexicon)

The NRC lexicon provides a richer emotional picture by categorizing words into specific emotions beyond just positive/negative.

```{r plot2, warning=FALSE}
nrc_sentiment <- got_tokens %>%
  inner_join(nrc, by = "word") %>%
  count(Season, sentiment) %>%
  mutate(Season = as.factor(Season))

nrc_plot <- nrc_sentiment %>%
  ggplot(aes(x = Season, y = n, fill = sentiment)) +
  geom_col() +
  facet_wrap(~sentiment, scales = "free_y") +
  scale_fill_viridis_d() +
  labs(title = "Emotion Distribution by Season", x = "Season", y = "Word Count", fill = "Emotion") +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", size = 12), legend.position = "bottom", strip.text = element_text(face = "bold", size = 9))

ggplotly(nrc_plot, tooltip = c("x", "y", "fill"))
```

### Lexicon Comparison Summary

- **Bing Lexicon** reveals an overall balance between positive and negative sentiment across the series, with notable variation by season.
- **NRC Lexicon** shows that emotions like positive, negative, trust are most prevalent in Game of Thrones dialogue, reflecting the show's dramatic and often dark narrative.

## Part 2: Sentiment Trajectory

How does emotional tone shift throughout the series? 

### Plot 3: Net Sentiment Over Time (by Episode)

This visualization tracks the cumulative sentiment trajectory across all episodes.

```{r plot3, warning=FALSE}
sentiment_trajectory <- got_tokens %>%
  inner_join(bing, by = "word") %>%
  mutate(sentiment_value = ifelse(sentiment == "positive", 1, -1)) %>%
  group_by(Season, Episode, episode_id) %>%
  summarise(net_sentiment = sum(sentiment_value), total_words = n(), sentiment_score = net_sentiment / total_words, .groups = 'drop') %>%
  arrange(Season, Episode) %>%
  mutate(episode_num = row_number())

trajectory_plot <- sentiment_trajectory %>%
  ggplot(aes(x = episode_num, y = sentiment_score, color = as.factor(Season), group = 1)) +
  geom_line(alpha = 0.6) +
  geom_point(alpha = 0.5, size = 2) +
  geom_smooth(se = TRUE, color = "black", alpha = 0.1, method = "loess", span = 0.3) +
  scale_color_viridis_d(name = "Season") +
  labs(title = "Sentiment Score Trajectory Across All Episodes", x = "Episode Number (Chronological)", y = "Sentiment Score (positive - negative) / total words", color = "Season") +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", size = 12))

ggplotly(trajectory_plot, tooltip = c("x", "y", "color"))
```

### Plot 4: Average Sentiment by Season

Which seasons are emotionally the most intense?

```{r plot4, warning=FALSE}
season_sentiment <- sentiment_trajectory %>%
  group_by(Season) %>%
  summarise(avg_sentiment = mean(sentiment_score), min_sentiment = min(sentiment_score), max_sentiment = max(sentiment_score), .groups = 'drop') %>%
  mutate(Season = as.factor(Season))

season_plot <- season_sentiment %>%
  ggplot(aes(x = Season, y = avg_sentiment, fill = avg_sentiment)) +
  geom_col(show.legend = FALSE) +
  geom_errorbar(aes(ymin = min_sentiment, ymax = max_sentiment), width = 0.3, alpha = 0.6) +
  scale_fill_gradient2(low = "red", mid = "gray", high = "green", midpoint = 0) +
  labs(title = "Average Sentiment Score by Season", subtitle = "Error bars show min-max range", x = "Season", y = "Average Sentiment Score") +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", size = 12), plot.subtitle = element_text(face = "italic",  size = 10))

ggplotly(season_plot, tooltip = c("x", "y"))
```

## Part 3: Contribution Analysis

Which specific words contribute the most to the overall sentiment? 

### Plot 5: Top Positive Word Contributions

Words with the highest positive contribution (frequency × positive sentiment value).

```{r plot5, warning=FALSE}
positive_contributions <- got_tokens %>%
  inner_join(bing, by = "word") %>%
  filter(sentiment == "positive") %>%
  count(word, sort = TRUE) %>%
  head(15) %>%
  mutate(word = reorder(word, n),
         contribution = n)

positive_plot <- positive_contributions %>%
  ggplot(aes(contribution, word, fill = contribution)) +
  geom_col(show.legend = FALSE) +
  scale_fill_gradient(low = "lightgreen", high = "darkgreen") +
  labs(title = "Top 15 Positive Words Contribution",
       x = "Frequency",
       y = NULL) +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", size = 12))

ggplotly(positive_plot, tooltip = c("x", "y"))
```

### Plot 6: Top Negative Word Contributions

Words with the highest negative contribution (frequency × negative sentiment value).

```{r plot6, warning=FALSE}
negative_contributions <- got_tokens %>%
  inner_join(bing, by = "word") %>%
  filter(sentiment == "negative") %>%
  count(word, sort = TRUE) %>%
  head(15) %>%
  mutate(word = reorder(word, n),
         contribution = n)

negative_plot <- negative_contributions %>%
  ggplot(aes(contribution, word, fill = contribution)) +
  geom_col(show.legend = FALSE) +
  scale_fill_gradient(low = "lightcoral", high = "darkred") +
  labs(title = "Top 15 Negative Words Contribution",
       x = "Frequency",
       y = NULL) +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", size = 12))

ggplotly(negative_plot, tooltip = c("x", "y"))
```

## Summary 

- The Bing lexicon captures a binary view of sentiment, which is useful for quick assessment. NRC provides emotional granularity that reveals specific emotional tones shown in the series.
- The series shows a most upward but mixed trajectory, withthe first season being most negative and last season being most positive. 
- The sentiment is driven primarily by recurring words that appear throughout the series, such as "good", "well", "right" for positive sentiment and "stark", "dead", "kill" for negative sentiment.