---
title: "My First Quarto Dashboard"
author: "Ian Cheung"
format: 
  dashboard:
    orientation: columns
embed-resources: true
---

```{r}
#| label: setup
#| message: false
library(tidyverse)
library(ggplot2)
library(gt)

```

## Column {width=60%}

### Research Question

How does vehicle class affect highway fuel economy, and which classes are the most efficient?

### Highway Fuel Economy by Vehicle Class

```{r}
#| label: fuel-graph
#| title: "Highway MPG Distribution"
ggplot(mpg, aes(x = reorder(class, hwy, FUN = median), y = hwy, fill = class)) +
  geom_boxplot(show.legend = FALSE) +
  labs(x = "Vehicle Class", y = "Highway MPG") +
  theme_minimal()

```

## Column {width=40%}

### Efficiency Leaders

```{r}
#| label: stats-value-box
#| component: valuebox
# Find the highest highway MPG in the dataset
max_hwy <- max(mpg$hwy)
list(
  icon = "fuel-pump",
  color = "success",
  value = paste(max_hwy, "MPG")
)

```

### Summary of Top Performers

```{r}
#| label: top-table
#| title: "Top 5 Most Efficient Models"
mpg %>%
  select(manufacturer, model, class, hwy) %>%
  arrange(desc(hwy)) %>%
  slice(1:5) %>%
  gt() %>%
  cols_label(
    manufacturer = "Make",
    model = "Model",
    class = "Type",
    hwy = "HWY MPG"
  )

```

### Report Summary

This dashboard analyzes the `mpg` dataset to determine the fuel efficiency for the vehicle classes. As shown in the boxplot, **subcompact**, **compact**, and **midsize** cars significantly outperform SUVs and pickups. The value box highlights the peak efficiency recorded in this dataset ($44$ MPG), while the table identifies the top 5 most efficient models.
