---
title: "EV Adoption"
subtitle: "California Energy Commission"
author: "Ian Cheung"
format:
  revealjs:
    code-folding: true
    echo: true
    embed-resources: true
    transition: slide
    background-transition: fade
---

## Introduction

This is the 2023 sales data to identify trends in adoption of EV. 

---

## State of the Counties {background-color="#eef6ff"}

```{r}
#| label: county-stats
#| code-line-numbers: "9-10"
library(dplyr)
ev_sales <- read.csv("ev-sales-cleaned.csv")

county_summary <- ev_sales %>%
  group_by(county) %>%
  summarise(total_sales = sum(sales)) %>%
  arrange(total_sales)

lowest_county <- filter(county_summary, total_sales == min(total_sales))
highest_county <- filter(county_summary, total_sales == max(total_sales))

```

* **Highest Adoption:** `r highest_county$county` (`r highest_county$total_sales` sales)
* **Lowest Adoption:** `r lowest_county$county` (`r lowest_county$total_sales` sales)

---

## Model Popularity {background-color="#fef8dbff"}

```{r}
#| label: model-stats
#| code-line-numbers: "6-7"
model_stats <- ev_sales %>%
  group_by(make, model) %>%
  summarise(total_model_sales = sum(sales)) %>%
  arrange(desc(total_model_sales))

top_model <- model_stats[1, ]
bottom_model <- model_stats[nrow(model_stats), ]

```

* **Most Popular:** `r top_model$make` `r top_model$model`
* **Least Popular:** `r bottom_model$make` `r bottom_model$model`