---
title: "California Crash Data Analysis (2023)"
subtitle: "Project 2"
author: "Ian Cheung"
format: 
  revealjs:
    slide-number: c
    embed-resources: true
    theme: simple
    transition: fade
---

```{r setup, include=FALSE}

library(tidyverse)
dat <- read_csv("crashes_2023.csv")
```

# Introduction

## About this Project

This project explores traffic crash data from the **Statewide Integrated Traffic Records System (SWITRS)**, maintained by the California Highway Patrol.

# 2.2 Overall Statistics

## Statewide Summary

```{r}
#| echo: true
#| code-fold: true

overall_stats <- dat |> 
  summarize(
    total_crashes = n(),
    total_injured = sum(NUMBER_INJURED, na.rm = TRUE),
    total_fatal = sum(NUMBER_KILLED, na.rm = TRUE)
  )

```

During the year 2023, California recorded the following:

* **Total Crashes:** `r pull(overall_stats, total_crashes)`
* **Total Injured Victims:** `r pull(overall_stats, total_injured)`
* **Total Fatal Victims:** `r pull(overall_stats, total_fatal)`

# 2.3 Distribution Tables

## Crashes per Month

```{r}
#| echo: true
#| code-fold: true

dat |> 
  group_by(MONTH) |> 
  summarize(count = n()) |> 
  mutate(percentage = round((count / sum(count)) * 100, 2)) |> 
  arrange(MONTH)

```

## Crashes per Day of Week

```{r}
#| echo: true
#| code-fold: true

dat |> 
  group_by(DAY_OF_WEEK) |> 
  summarize(count = n()) |> 
  mutate(percentage = round((count / sum(count)) * 100, 2)) |> 
  arrange(DAY_OF_WEEK)

```

## Crashes per Hour

```{r}
#| echo: true
#| code-fold: true

dat |> 
  group_by(HOUR) |> 
  summarize(count = n()) |> 
  mutate(percentage = round((count / sum(count)) * 100, 2)) |> 
  arrange(HOUR)

```

# 2.4 Ranking Tables

## Top-10 PCF Violations

```{r}
#| echo: true
#| code-fold: true

dat |> 
  count(PCF_VIOL_CATEGORY, name = "count") |> 
  slice_max(order_by = count, n = 10) |> 
  arrange(desc(count))

```

## Types of Collision

```{r}
#| echo: true
#| code-fold: true

dat |> 
  count(TYPE_OF_COLLISION, name = "count") |> 
  arrange(desc(count))

```

## Top-8 Party At Fault

```{r}
#| echo: true
#| code-fold: true

dat |> 
  count(STWD_VEHTYPE_AT_FAULT, name = "count") |> 
  slice_max(order_by = count, n = 8) |> 
  arrange(desc(count))

```

# 2.5 Pedestrian Involvement

## Pedestrian Totals

```{r}
#| echo: true
#| code-fold: true

ped_stats <- dat |> 
  filter(PEDESTRIAN_ACCIDENT == "yes") |> 
  summarize(
    total_accidents = n(),
    total_injured = sum(COUNT_PED_INJURED, na.rm = TRUE),
    total_killed = sum(COUNT_PED_KILLED, na.rm = TRUE)
  )

```

* **Accidents involving pedestrians:** `r pull(ped_stats, total_accidents)` 
* **Pedestrians injured:** `r pull(ped_stats, total_injured)` 
* **Pedestrians killed:** `r pull(ped_stats, total_killed)` 

## Pedestrian Rankings 1
### Top 5 Violation Category

```{r}
#| echo: true
#| code-fold: true

dat |> 
  filter(PEDESTRIAN_ACCIDENT == "yes") |> 
  count(PCF_VIOL_CATEGORY) |> 
  slice_max(order_by = n, n = 5)
```
## Pedestrian Rankings 2
### Top 5 Collision Types
```{r ped-rankings2}
#| echo: true
#| code-fold: true
dat |> 
  filter(PEDESTRIAN_ACCIDENT == "yes") |> 
  count(TYPE_OF_COLLISION) |> 
  slice_max(order_by = n, n = 5)
```
## Pedestrian Rankings 3
### Top 5 Vehicle At Fault
```{r}
#| echo: true
#| code-fold: true
dat |> 
  filter(PEDESTRIAN_ACCIDENT == "yes") |> 
  count(STWD_VEHTYPE_AT_FAULT) |> 
  slice_max(order_by = n, n = 5)

```

# 2.6 Bicyclist Involvement

## Bicyclist Totals

```{r}
#| echo: true
#| code-fold: true

bike_stats <- dat |> 
  filter(BICYCLE_ACCIDENT == "yes") |> 
  summarize(
    total_accidents = n(),
    total_injured = sum(COUNT_BICYCLIST_INJURED, na.rm = TRUE),
    total_killed = sum(COUNT_BICYCLIST_KILLED, na.rm = TRUE)
  )

```

* **Accidents involving bicyclists:** `r pull(bike_stats, total_accidents)` 
* **Bicyclists injured:** `r pull(bike_stats, total_injured)` 
* **Bicyclists killed:** `r pull(bike_stats, total_killed)` 


## Bicycle Rankings 1
### Top 5 Violation Category

```{r}
#| echo: true
#| code-fold: true

dat |> 
  filter(BICYCLE_ACCIDENT == "yes") |> 
  count(PCF_VIOL_CATEGORY) |> 
  slice_max(order_by = n, n = 5)
```
## Bicycle Rankings 2
### Top 5 Collision Types
```{r}
#| echo: true
#| code-fold: true
dat |> 
  filter(BICYCLE_ACCIDENT == "yes") |> 
  count(TYPE_OF_COLLISION) |> 
  slice_max(order_by = n, n = 5)
```
## Bicycle Rankings 3
### Top 5 Vehicle At Fault
```{r}
#| echo: true
#| code-fold: true
dat |> 
  filter(BICYCLE_ACCIDENT == "yes") |> 
  count(STWD_VEHTYPE_AT_FAULT) |> 
  slice_max(order_by = n, n = 5)

```

# 2.7 County Exploration

## High-Crash Counties

**Research Question:** Which five California counties reported the highest number of crashes in 2023?

```{r}
#| echo: true
#| code-fold: true

dat |> 
  count(COUNTY, name = "total_crashes") |> 
  slice_max(order_by = total_crashes, n = 5) |> 
  mutate(percentage = round((total_crashes / nrow(dat)) * 100, 2)) |>
  arrange(desc(total_crashes))

```
